\myTitle

### 2015-03-06 tex/abstract.tex
Added new figures, edited each section, made all figures in color and greyscale

      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
interpreted as an estimation procedure using  
a model that is essentially identical to the {\sc structure} 
\citep{Pritchardetal2000} model with no admixture and non-correlated 
allele frequency priors. The computational routine in {\sc flock} is 
equivalent to searching for the maximum-{\em a-posteriori}
estimate of this restricted {\sc structure} model via a simulated 
annealing algorithm with an extreme cooling 
schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  By analysis of a large, real data set we
confirm that {\sc flock} yields results similar to {\sc structure} with the no-admxixture
model and uncorrelated allele frequency prior with more variability 
from run to run. Interpreting {\sc flock}'s algorithm as a special case of the model in 
{\sc structure} should aid in  understanding the program's output and behavior.

>>>>

      We show that the the algorithm in the program {\sc flock} \citep{Duc&Tur2009} can be
interpreted as an estimation procedure using  
a model that is essentially identical to the {\sc structure} 
\citep{Pritchardetal2000} model with no admixture and non-correlated 
allele frequency priors. The computational routine in {\sc flock} is 
equivalent to searching for the maximum-{\em a-posteriori}
estimate of this restricted {\sc structure} model via a simulated 
annealing algorithm with an extreme cooling 
schedule (namely, the exponent on the objective function~$\rightarrow \infty$).  We 
demonstrate the similarities between the two programs in a two step approach. First,
we modified the source code of  {\sc structure} to resemble {\sc flock} producing
the program {\sc flockture}. With simulated data we confirm similar behavior between 
{\sc flock}  and {\sc flockture} by assing the programs' performance
to assign individuals back to their correct population with a 
0-1 loss function (ie. a loss of 1 is incurred if an individual is assinged 
to an incorrect group). Second, we simulated many datasets under varying 
levels of population differentiation for both microsatllite and SNP genoyptes. We show that
{\sc flockture} yields results similar to {\sc structure} although with much larger 
variability from run to run. {\sc flockture} did perform better than {\sc structure} 
when genotypes were composed of SNPs and differentiation was moderate 
($F_{ST}=0.022-0.032$). When differentiation was low, {\sc structure} had a lower loss
for both marker types. Interpreting {\sc flock}'s algorithm as a special case of the model in 
{\sc structure} should aid in understanding the program's output and behavior.

### 2015-03-06 tex/main-body-text.tex
Added new figures, edited each section, made all figures in color and greyscale

Being able to interpret {\sc flock} in this fashion should help users
interpret output from {\sc flock} and predict when it might and might not
give substantially different results than {\sc structure}. Below we describe
the correspondence between the programs in more detail, and then, for illustration, compare
the results obtained from {\sc flock} and {\sc structure} on a large, real data
set.

@@

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} for estimating the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur2012}. In doing so, 
{\sc flock} relies on individuals with like genotypes clustering together 
through an iterative process. It stands to reason then, a reasonable way 
to gauge how well the algorithm accomplishes the main objective 
of clustering individuals into their correct groups is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does assuming the correct number of populations. 
The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel precludes a rigorous evaluation of its behavior due 
to the inability to efficiently batch process. Consequenty, we adapted 
the source code from {\sc structure} as describe above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrated that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulated and analyzed datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A-C) model, the no-admixture and correlated 
allele frequency (NA-C) model, and the no-admixture and non-correlated 
allele frequency (NA-NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .

@@

To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
were composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
populations. Both programs were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as
 defined by a zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

@@

Run times for both programs were relatively short, albeit in the case of {\sc flock} 
non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
general increase as population structure became more subtle, however {\sc flock} did not 
demonstrate a linear increase. % this is surprising to me as it is coin the same thing irrespective of FST
% What I assume is happening is that as FST increases you have a lot more individuals being shuffled around
% at later iterations leading to increased process times. Seems reasonable.

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA-NC and NA-C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
resulting from the assignment of individuals from two groups into a single cluster while 
individuals from another group were separated into two clusters
with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct}  A). This phenomena 
(only present in a single microsatellite dataset) disappears as population differentiation 
declines. As the subtlety of genetic structure 
increased loss for both programs increases and differences in the performance of
each model with different marker types emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increases and certain models do marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ drops 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
 {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number and length of plateaus {\sc flockture} encountered quickly 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. The other eight 
replicate runs yielded plateau records varying in length from two to eight all of which had at 
least one plateau larger than thirteen. When $F_{ST}$ was larger than 0.0337 no plateau record
contained at least one plateau of length six or greater. No plateaus were observed for any run when 
$F_{ST}$ was larger than 0.02187. Results from the SNP datasets
were concordant. No run of {\sc flockture} resulted in a single plateau of length greater than six.
Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets 
graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
likelihood were suggestive of population structure.

Plots of the normalized likelihood values from {\sc flockture} can be useful in discerning the presence of 
structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was indistinguisable
and results were highly concordant with those of {\sc structure}. As population differentiation degrades 
and plateaus are not observed, the run with the highest likelihood from {\sc flockture} still yields
similar clusters as {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases the normalized likelihood
values for assignment remain high in contrast to the $q_i$ values in each of the three models of {\sc structure}.

  \begin{figure*}
\centering
\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatloss.pdf}%Need to make a greyscale version
  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets were simulated with varying levels of population differentation. A coalescent based program
was used to generate 15 loci microsatellite genotypes for 2000 individuals comprising five populations.}
  \label{fig:uSatLoss}
\end{figure*}

 \begin{figure*}
\centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%Need to make a greyscale version
  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets were simulated with varying levels of population differentation. A coalescent based program
was used to generate 96 loci single nucleotide polymorphism (SNP) genotypes for 2000 
individuals comprising five populations.}
  \label{fig:SNPloss}
\end{figure*}

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}
  \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values from
  {\sc flockture} made with the program {\sc distruct}. Each horizontal plot represents one of nine 
 runs of the programs  at $K$ = 5. Three models were run in structure: the admixture and correlated allele 
  frequency (A-C) model, 
the no-admixture and correlated allele frequency (NA-C) model, 
and the no-admixture and non-correlated allele frequency (NA-NC) model. Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became somewhat inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence and also
the large size of the search space over possible partitions. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
By having a large number of individuals
that can be arranged into 5 groups defines a large partion space.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This could act by increasing the number of
local peaks, leading {\sc flockture} settle on many different solutions. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flockture}.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods), 
their comparision was based on setting {\sc flock} to perform 50 runs and 20 iterations for each 
$K$ starting at 2 and ending when one of the stopping conditions was met, and computing times for that analysis 
were compared to ten iterations of the admixture and correlated allele frequency model in {\sc structure} using 
a 50,000 sweep burn-in followed by a 200,000 sweep sample 
from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. It is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
{\sc flock} were to be implemented in a coding language rather than a VBA routine in microsoft excel
it would be approximately 200 times faster.

>>>>

Being able to interpret {\sc flock} in this fashion should help users
interpret output from {\sc flock} and predict when it might and might not
give substantially different results than {\sc structure}. Below we describe
the correspondence between the programs in more detail, and then, for illustration,
create the progam {\sc flockture} by adapting the code of {\sc structure} to 
use a simulated annealing approach. We then compare results from large simulated 
datasets with varying degrees of population differentiation.

@@

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} for estimating the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur2012}. In doing so, 
{\sc flock} relies on individuals with like genotypes clustering together 
through an iterative process. It stands to reason then, a reasonable way 
to gauge how well the algorithm accomplishes the main objective 
of clustering individuals into their correct groups is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does assuming the correct number of populations. The 
algorithme performance can then be measured with a zero-one 
loss function (where a loss of 1 is incurred if an individual is misassigned).

The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel precludes a rigorous evaluation of its behavior due 
to the inability to efficiently batch process. Consequenty, we adapted 
the source code from {\sc structure} as describe above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrated that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulated and analyzed datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A.C) model, the no-admixture and correlated 
allele frequency (NA.C) model, and the no-admixture and non-correlated 
allele frequency (NA.NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .

@@

To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
were composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
populations. Both programs were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as
 defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

@@

Run times for both programs were relatively short, albeit in the case of {\sc flock} 
non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
general increase as population structure became more subtle, however {\sc flock} did not 
demonstrate a linear increase.

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA.NC and NA.C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
resulting from the assignment of individuals from two groups into a single cluster while 
individuals from another group were separated into two clusters
with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct}  A). This phenomena 
(only present in a single microsatellite dataset) disappeared as population differentiation 
declined. As the subtlety of genetic structure 
increased loss for both programs increased and differences in the performance of
each model with different marker types emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increased and certain models did marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ droped 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
 {\sc flockture} did, however, have consistenlty smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016 {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number and length of plateaus {\sc flockture} encountered quickly 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets
only one run at $F_{ST} = 0.0339$ produced a single plateau of length greater than six. The other eight 
replicate runs yielded plateau records varying in length from two to eight, all of which had at 
least one plateau larger than thirteen. When $F_{ST}$ was larger than 0.0337 no plateau record
contained a plateau of length six or greater. No plateaus were observed for any run when 
$F_{ST}$ was larger than 0.02187. Results from the SNP datasets
were concordant. No run of {\sc flockture} resulted in a single plateau of length greater than six.
Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets 
graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
likelihood were suggestive of population structure.

Plots of the normalized likelihood values from {\sc flockture} can be useful in discerning the presence of 
structure.  In situations where only a single run of {\sc flockture} produced a plateau of length 2 
(Fig.~\ref{fig:FvSdistruct} B) the variation in assignment of individuals among runs was qualitatively indistinguisable
and results were highly concordant with those of {\sc structure}. As population differentiation degrades 
and plateaus are not observed, the run with the highest likelihood from {\sc flockture} still yields
similar clusters as {\sc structure} but with more variation among runs (Fig.~\ref{fig:FvSdistruct} C \& D).
As pairwise $F_{ST}$ decreases and the variation among {\sc flockture} runs increases the normalized likelihood
values for assignment remain high in contrast to the $q_i$ values in each of the three models of {\sc structure}.

  \begin{figure*}
\centering
\includegraphics[width=.9\linewidth]{images/Figures-Pat/uSatLoss.pdf}%uSatloss_BW.pdf is a greyscale version
  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets, composed of 2,000 individuals with 15 microsatellite genotypes comprising
5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
of both programs. }
  \label{fig:uSatLoss}
\end{figure*}

 \begin{figure*}
\centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/SNPloss.pdf}%SNPloss_BW.pdf is a greyscale version
  \caption{Zero-one loss for nine runs of {\sc flockture} and the admixture and correlated allele 
  frequency (A.C) model, the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model in {\sc structure}. Forty-eight 
datasets, composed of 2,000 individuals with 96 SNP genotypes comprising 
5 populations of sizes 250, 250, 400, 550, and 550, were simulated to evaluate the performance
of both programs. }
  \label{fig:SNPloss}
\end{figure*}

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
  \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values from
  {\sc flockture} made with the program {\sc distruct}. Each horizontal plot represents one of nine 
 runs of the programs  at $K$ = 5. Three models were run in structure: the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model. Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became somewhat inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence and also
the large size of the search space over possible partitions. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
By having a large number of individuals
that can be arranged into 5 groups defines a large partion space.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This could act by increasing the number of
local peaks, leading {\sc flockture} to settle on many different solutions. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our results suggest this is a bigger problem for {\sc flockture}.
In species characterized by large subpopulations and low divergence, as is the case with many 
marine fish, the use of {\sc flock} should probably be avoided.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods), 
their comparision was based on setting {\sc flock} to perform 50 runs and 20 iterations for each 
$K$ starting at 2 and ending when one of the stopping conditions was met, and computing times for that analysis 
were compared to ten iterations of the admixture and correlated allele frequency model in {\sc structure} using 
a 50,000 sweep burn-in followed by a 200,000 sweep sample 
from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. It is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
{\sc flock} were to be implemented in a coding language rather than a VBA routine in Microsoft Excel
it would be approximately 200 times faster.
